TacticToe: Learning the Game of Tactics

### 2018-02-08 tactictoe_jv.tex
minor

\lstdefinelanguage{SMLSmall}%
{columns=fullflexible,
  keywords={THEN,THENL,val,open,store_thm,fun,fn,let,in,end},%
  frame=none,
  sensitive=true,
  keywordstyle=\fontfamily{lmss}\scriptsize\selectfont,%
  basicstyle=\fontfamily{pcr}\small\selectfont,%
  stringstyle=\tt,
  morestring=[b]",
  literate={\ }{$\mkern6mu$}1%
   {=}{{\tt\raisebox{-.15mm}{=}}}1%
   {[}{{\tt\raisebox{-.15mm}{[}}}1%
   {]}{{\tt\raisebox{-.15mm}{]}}}1%
   {->}{{$\rightarrow$}}1%
   {==>}{{$\Rightarrow$}}1%
   {wedge}{{$\wedge$}}1%
   {<=}{{$\leq$}}1%
   {>=}{{$\geq$}}1%
   {'a}{{$\alpha$}}1%
   {'b}{{$\beta$}}1%
   {ldots}{$\ldots$}1%
   {!}{$\forall$}1%
   {fff}{{DB.fetch}}1%
   {``}{\hspace{-.5mm}`\hspace{-1mm}`}1%
  }
\lstset{language=SML}
\begin{document}
\maketitle

@@

We bridge this gap by implementing our tactical prover TacticToe on top
of the HOL4 ITP.
TacticToe first learns from many proof scripts
the relevance of tactics in particular proof situations.
This learned knowledge is then used in a Monte Carlo tree search algorithm that explores the most promising tactic-level proof paths.
On a single CPU, with a time limit of 60 seconds, TacticToe proves 66.2\%
of the 7164 theorems in HOL4's standard library whereas
Eprover solves 34.5\%. The success rate rises to 69.0\% by combining the
results of both provers.
\end{abstract}

@@

\begin{definition}(search tree)\\
A search tree $\mathfrak{T}$ is a sixtuple
$(\mathbb{T},\mathbb{G},\mathbb{A},T,G,A)$
that respects the following conditions:
\begin{itemize}
\item $\mathbb{T}$ is a set of tactics, $\mathbb{G}$ is a set of goals
 and $\mathbb{A}$ is a set of nodes.
\item $T$ is a total function from $\mathbb{G}$ to $\mathcal{P}(\mathbb{T})$. It takes a
goal and return the set of tactics already applied to this goal.
\item $G$ is a total function from $\mathbb{A}$ to $\mathcal{P}(\mathbb{G})$. It takes a
node and returns the set of goals in this node.

\item $A$ is a partial function from $\mathbb{G} \times \mathbb{T}$ to
$\mathbb{A}$. It takes a goal and a tactic and returns the node produced by the
application of the tactic to the goal. Its support $\mathit{Support}(A)$ is by
definition $\lbrace (g,t) \in
\mathbb{G} \times \mathbb{T}\ |\ t \in T(g)
\rbrace$.
\item If $(g,t) \in \mathit{Support}(A)$, then $t(g) = G(A(g,t))$. This is the
main condition. The result of the application of a tactic $t$
to a goal $g$ is exactly the list of goals contained in the node produced by
$t$ from
$g$.
\end{itemize}

@@

If no explicit order is given, we assume that any set is equipped with an
arbitrary order. In this way, we can access the $n^{th}$ member of a set $s$ by
writing $s_n$. In Figure~\ref{fig:choice}, $(G(a_j))_i$ is written a $g_j^i$.

@@

Starting from the bottom parent node, we can construct children of a goal of
this node by choosing a goal $g_i^0$ on this node and choosing tactics
$t_1,t_2,\ldots$ to
apply to this goal. Here the result produces new nodes each containing a list
of goals to be proven called obligations. In each node of the proof tree,
a goal can either be
solved, open or pending (see Definition~\ref{def:solved}).

@@

\begin{definition}\label{def:desc}(Children, descendant, active node)\\
We now define what are the children and the descendants of a node $a$. The
children of a relative to a goal $g$ in $G(a)$
is the set of nodes produced by tactics applied to $g$:
  \[\mathit{Children}(a,g) = \lbrace A(g,t)\in \mathbb{A}\ |\ t \in T(g) \rbrace \]

If $g$ is the open goal of $a$, then we simply call children of a the
children of $a$ relative to $g$. We define:\\

Extended children of $a$ by:
  \[\mathit{ExtChildren}(a) = \bigcup_{g \in G(a)} \mathit{Children}(a,g) \]

Descendants of $a$ by:
\begin{align*}
\mathit{Descendant}^{0}(a) &=_{def} \lbrace a \rbrace \\
\mathit{Descendant}^{n+1}(a) &=_{def} \bigcup_{a' \in \mathit{Descendant}^{n}(a)}
\mathit{ExtChildren}(a') \\
\mathit{Descendant}(a) &=_{def} \bigcup_{n \in \mathbb{N}} \mathit{Descendant}^n(a)\\
\end{align*}

%Descendants of $a$ through $g$ by:
%\[Descendant(a,g) =_{def}
%  \lbrace a \rbrace \cup \bigcup_{a' \in Children(a,g)} Descendant(a')\]

@@

Through node extension steps the search tree grows and the number of paths to
explore increases. To decide which node to extend next, a evolving value for
each node is given in Definition~\ref{def:value}. The algorithm that performs
node selection starts from the root of the search tree. X, it repeatedly choses
the children with the highest values. Depending on the widening policy, it can
stop at any chosen node on during its descent through the tree. We now give a
more precise description of
the algorithm with the following pseudo-code:

\todoi{Replace alltt environment by something prettier}
\begin{alltt}
ChosenNode = Root(Tree);
while true do
  if Children(ChosenNode) = \(\emptyset\) then break;
  (BestChild, BestValue) = CompareChildren (ChosenNode);
  if WideningPolicy (ChosenNode) \(\geq\) BestValue then break;
  ChosenNode = BestChild
end;
return ChosenNode;
\end{alltt}

@@

\begin{table}[h!]
\centering\ra{1.3}
\small
\begin{tabular}{lc}
\toprule
  & Solved (60s) \\
\midrule
   \eprover   & 2472 (34.5\%)\\
   \tactictoe & 4741 (66.2\%)\\
\midrule
   Total  & 4946 (69.0\%)\\
\bottomrule
\end{tabular}
\caption{Evaluation on 7164 top-level theorems of the \holfour standard library
\label{tab:_param}}
\end{table}

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=north east, at={(0.9,0.9)}},
  width=0.5*\textwidth,
  height=0.5*\textwidth,
  ymin=0, ymax=3000,
  xmin=0, xmax=10,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=length, y=proofs] {data/tactictoe_proof_length};
\addplot table[x=length, y=proofs] {data/original_proof_length};
\end{axis}
\end{tikzpicture}
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=north east, at={(0.9,0.9)}},
  width=0.5*\textwidth,
  height=0.5*\textwidth,
  ymin=0, ymax=120,
  xmin=10, xmax=50,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=length, y=proofs] {data/tactictoe_proof_length};
\addplot table[x=length, y=proofs] {data/original_proof_length};
\legend{\tactictoe proofs,human proofs}
\end{axis}
\end{tikzpicture}
\caption{\label{fig:distrib}Number of proofs (y axis) with respect to their 
length (x axis) in 
number of tactic units. Comparison between the set of \tactictoe proofs and the 
set of human proofs.}
\end{figure}

@@

\todoi{Is copied and modified}
We proposed a new proof assistant automation technique which combines
tactic-based proof search with machine learning tactic prediction
Its implementation,
\tactictoe, achieves an overall performance of 66.2\% theorems on 7164
theorems
of the \holfour standard library surpassing \eprover with
auto-schedule. Its
effectiveness is especially visible on
theories which use inductive data structures, specialized decision procedures,
and custom built simplification sets.
Thanks to the learning abilities of \tactictoe, the generated proof scripts
usually reveal the high-level structure of the proof. %For these reasons,
We therefore believe that predicting ITP tactics based on the current goal
features is a very reasonable approach to automatically guiding proof search,
and that accurate predictions can be obtained by learning from the knowledge
available in today's large formal proof corpora.
%And it seems to get us
%closer to the way how humans learn to do theorem proving.

>>>>

\lstdefinelanguage{SMLSmall}%
{columns=fullflexible,
  keywords={THEN,THENL,val,open,store_thm,fun,fn,let,in,end,true,
  while,do,if,then,else,break,return,;},%
  frame=none,
  sensitive=true,
  keywordstyle=\fontfamily{lmss}\scriptsize\selectfont,%
  basicstyle=\fontfamily{pcr}\small\selectfont,%
  stringstyle=\tt,
  morestring=[b]",
  literate={\ }{$\mkern6mu$}1%
   {=}{{\tt\raisebox{-.15mm}{=}}}1%
   {[}{{\tt\raisebox{-.15mm}{[}}}1%
   {]}{{\tt\raisebox{-.15mm}{]}}}1%
   {->}{{$\rightarrow$}}1%
   {==>}{{$\Rightarrow$}}1%
   {wedge}{{$\wedge$}}1%
   {<=}{{$\leq$}}1%
   {>=}{{$\geq$}}1%
   {'a}{{$\alpha$}}1%
   {'b}{{$\beta$}}1%
   {ldots}{$\ldots$}1%
   {emptyset}{$\emptyset$}1%
   {!}{$\forall$}1%
   {fff}{{DB.fetch}}1%
   {``}{\hspace{-.5mm}`\hspace{-1mm}`}1%
  }
\lstset{language=SML}
\begin{document}
\maketitle

@@

We bridge this gap by implementing our tactical prover TacticToe on top
of the HOL4 ITP.
TacticToe first learns from many proof scripts
the relevance of tactics in particular proof situations.
This learned knowledge is then used in a Monte Carlo tree search algorithm that explores the most promising tactic-level proof paths.
On a single CPU, with a time limit of 60 seconds, TacticToe proves 66.4\%
of the 7164 theorems in HOL4's standard library whereas
Eprover solves 34.5\%. The success rate rises to 69.0\% by combining the
results of both provers.
\end{abstract}

@@

\begin{definition}(search tree)\\
A search tree $\mathfrak{T}$ is a sixtuple
$(\mathbb{T},\mathbb{G},\mathbb{A},T,G,A)$
that respects the following conditions:
\begin{itemize}
\item $\mathbb{T}$ is a set of tactics, $\mathbb{G}$ is a set of goals
 and $\mathbb{A}$ is a set of nodes representing lists of goals.
\item $T$ is a total function from $\mathbb{G}$ to $\mathcal{P}(\mathbb{T})$. It takes a
goal and return the set of tactics already applied to this goal.
\item $G$ is a total function from $\mathbb{A}$ to $\mathcal{P}(\mathbb{G})$. It takes a
node and returns the list of goals of this node.
\item $A$ is a partial function from $\mathbb{G} \times \mathbb{T}$ to
$\mathbb{A}$. It takes a goal and a tactic and returns the node produced by the
application of the tactic to the goal. Its support $\mathit{Support}(A)$ is by
definition $\lbrace (g,t) \in
\mathbb{G} \times \mathbb{T}\ |\ t \in T(g)
\rbrace$.
\item If $(g,t) \in \mathit{Support}(A)$, then $t(g) = G(A(g,t))$. This is the
main condition. The result of the application of a tactic $t$
to a goal $g$ is exactly the list of goals contained in the node produced by
$t$ from $g$.
\end{itemize}

@@

If no explicit order is given, we assume that any set is equipped with an
arbitrary order. In this way, we can access the $n^{th}$ member of a set $s$ by
writing $s_n$. In Figure~\ref{fig:choice}, $(G(a_j))_i$ is written as $g_j^i$.

@@

Starting from the parent node, we can construct children of a goal of
this node by choosing a goal $g_i^0$ on this node and choosing tactics
$t_1,t_2,\ldots$ to
apply to this goal. Here the result produces new nodes each containing a list
of goals to be proven called obligations. In each node of the proof tree,
a goal can either be
solved, open or pending (see Definition~\ref{def:solved}).

@@

\begin{definition}\label{def:desc}(Children, descendant, active node)\\
We now define what are the children and the descendants of a node $a$. The
children of a relative to a goal $g$ in $G(a)$
is the set of nodes produced by tactics applied to $g$:
\[\mathit{Children}(g) = \lbrace A(g,t)\in \mathbb{A}\ |\ t \in T(g) \rbrace\]

We define the children of a node $a$ by:
\[\mathit{Children}(a) = \bigcup_{g \in G(a)} \mathit{Children}(g) \]

And the descendants of $a$ by:
\begin{align*}
\mathit{Descendant}^{0}(a) &=_{def} \lbrace a \rbrace \\
\mathit{Descendant}^{n+1}(a) &=_{def} \bigcup_{a' \in \mathit{Descendant}^{n}(a)}
\mathit{Children}(a') \\
\mathit{Descendant}(a) &=_{def} \bigcup_{n \in \mathbb{N}} \mathit{Descendant}^n(a)\\
\end{align*}

@@

Through node extension steps the search tree grows and the number of paths to
explore increases. To decide which node to extend next, a evolving value for
each node is given in Definition~\ref{def:value}. The algorithm that performs
node selection starts from the root of the search tree. From its current node,
it choses among the children of the open goal the one with the highest value. 
If its value is higher than the widening policy, then the selected child 
becomes the current node, otherwise the final result of node selection is the 
current node. The node selection algorithm can be written as the following 
pseudo-code:

\begin{lstlisting}[language=SMLSmall]
CurrentNode = Root(Tree);
while true do
  if Children(CurrentNode) = emptyset then break;
  (BestChild, BestValue) = CompareChildren (CurrentNode);
  if WideningPolicy (ChosenNode) >= BestValue then break;
  CurrentNode = BestChild
end;
return CurrentNode;
\end{lstlisting}

@@

\begin{table}[h!]
\centering\ra{1.3}
\small
\begin{tabular}{lc}
\toprule
  & Solved (60s) \\
\midrule
   \eprover   & 2472 (34.5\%)\\
   \tactictoe & 4760 (66.4\%)\\
\midrule
   Total  & 4946 (69.0\%)\\
\bottomrule
\end{tabular}
\caption{Evaluation on 7164 top-level theorems of the \holfour standard library
\label{tab:_param}}
\end{table}

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=north east, at={(0.9,0.9)}},
  width=0.5*\textwidth,
  height=0.5*\textwidth,
  ymin=0, ymax=3000,
  xmin=0, xmax=10,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=length, y=proofs] {data/tactictoe_proof_length};
\addplot table[x=length, y=proofs] {data/original_proof_length};
\end{axis}
\end{tikzpicture}
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=north east, at={(0.9,0.9)}},
  width=0.5*\textwidth,
  height=0.5*\textwidth,
  ymin=0, ymax=120,
  xmin=10, xmax=50,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=length, y=proofs] {data/tactictoe_proof_length};
\addplot table[x=length, y=proofs] {data/original_proof_length};
\legend{\tactictoe proofs,human proofs}
\end{axis}
\end{tikzpicture}
\caption{\label{fig:distrib}Number of proof scripts with $x$ tactic units.}
\end{figure}

@@

\todoi{Is copied and modified}
We proposed a new proof assistant automation technique which combines
tactic-based proof search with machine learning tactic prediction
Its implementation,
\tactictoe, achieves an overall performance of 66.4\% theorems on 7164
theorems
of the \holfour standard library surpassing \eprover with
auto-schedule. Its
effectiveness is especially visible on
theories which use inductive data structures, specialized decision procedures,
and custom built simplification sets.
Thanks to the learning abilities of \tactictoe, the generated proof scripts
usually reveal the high-level structure of the proof. %For these reasons,
We therefore believe that predicting ITP tactics based on the current goal
features is a very reasonable approach to automatically guiding proof search,
and that accurate predictions can be obtained by learning from the knowledge
available in today's large formal proof corpora.
%And it seems to get us
%closer to the way how humans learn to do theorem proving.
